---
title: "Introduction to academiadates"
author: "Steven Carter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tsibble}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

I've created the `{academiadates}` package to extend support of the [`tidyverts`](https://tidyverts.org/) ecosystem handling a specific pattern of dates.
The pattern of dates is academic terms, hence the name.
Most institutions that I'm aware of use three terms per calendar year: spring, summer, and fall.

The `{tsibble}` package provides a new data frame class, the [`tsibble`](https://tsibble.tidyverts.org/reference/tsibble.html), and a handful of classes for various time intervals.
Those are covered on the [`{tsibble}` home page](https://tsibble.tidyverts.org/index.html).
This package provides a new class, `yeartrimester`, which is suited to handle three terms in a similar way as [`yearquarter`](https://tsibble.tidyverts.org/reference/year-quarter.html).

## Creation

```{r message=FALSE}
library(academiadates)
library(tsibble)
library(tsibbledata) # an example dataset
library(dplyr)
```

We have the following dataset from `{tsibbledata}` which is monthly data with counts of Australian livestock used for food.

```{r}
aus_livestock
```

Notice how the first month is July.
We can set the fiscal or academic year to start the seventh month for the sake of having a whole first year.

As an example we can aggregate this to quarters and trimesters.
We have to convert to a regular `tibble` to make regrouping the data under the new time index easier.

```{r}
# quarters
aus_q <- aus_livestock |> 
    as_tibble() |> 
    mutate(Quarter = yearquarter(Month, fiscal_start = 7)) |> 
    summarize(
        Count = sum(Count),
        .by   = c(Quarter, Animal, State)
    ) |> 
    as_tsibble(index = Quarter, key = c(Animal, State))

# trimesters
aus_t <- aus_livestock |> 
    as_tibble() |>
    mutate(Trimester = yeartrimester(Month, academic_start = 7)) |>
    summarize(
        Count = sum(Count),
        .by   = c(Trimester, Animal, State)
    ) |> 
    as_tsibble(index = Trimester, key = c(Animal, State))
```

```{r}
aus_q

aus_t
```

If you have an unusual form of date, such as `Fall 2025`, you can convert the column to a date type, `2025-09-01`, and then convert to `yeartrimester` with `yeartrimester()`.
(I intend to add a conversion from a term name with calendar year directly to a `yeartrimester` object.)

## Plotting

The built-in plotting functionality is limited.
The plotting functionality for the `tidyverts` ecosystem is partly from the `{feasts}` package: [Graphics](https://feasts.tidyverts.org/reference/index.html#graphics).
Some of those functions depend on the class of the index.
This means that, without porting those functions to accommodate the `yeartrimester` class, a portion of the functionality will not work.

(I hope to do that in the future, but the functionality can be recreated manually for the time being.)

```{r}
library(ggplot2)

aus_t |> 
    filter(
        Animal == "Bulls, bullocks and steers",
        Trimester >= yeartrimester('2012 T1', academic_start = 7)
    ) |> 
    ggplot(
        aes(
            x = Trimester,
            y = Count,
            color = State
        )
    ) +
    geom_line()
```

> There is a bug in `{tsibble}` when using `yearquarter` with a `fiscal_start`.
> `yearquarter_trans` does not account for that offset, so the `fiscal_start` is dropped through transforming to a date and then inverting the transform.
> 
> I've worked around this by adding in the additional months to it when transforming to a date.
> The re-transformed `yeartrimester` does not have an offset, but the offset is built-in.
